Copulas and their potential for ecology

### 2019-08-23 Paper.Rmd
starting to address referee comments from Ecol Mongr, only inserted comments into Paper.Rmd so far, but about to start actually editing

All branches of ecology 
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have profound significance, as discussed below, but is not captured by 
correlation. 
Monotonic transformations can alter the Pearson, but not the Spearman
correlation (Appendix \ref{SM-pedagog1_details}, 
Fig. \ref{SM-fig_transf_pearson_spearman}). Although correlation
is only one way to 
study relationships between variables, other common metrics also 
provide partial information.

@@

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*, which we here 
introduce. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure.
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A main benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for Q1 below).

The goal of this paper is to explore the potential for 
applications of copulas in ecology. 
Our results provide support for the hypothesis that environmental, ecological and evolutionary processes 
generate and transmit nontrivial copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. The work we present here has begun to test this
hypothesis, and we hope our results stimulate further
evaluation of the hypothesis by the research community.

@@

One way we may expect, *a priori*, a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association is visible in the copula
approach, but may not be revealed by standard techniques.

@@

A third reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
would suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F suggests that Sp2
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between 
variables, and this 
information can then help support or refute alternative causal hypotheses
about which species is driving the dynamics of the other.

@@

>>>>

All branches of ecology 
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have significance, as discussed below, but is not captured by 
correlation. 
Monotonic transformations can alter the Pearson, but not the Spearman
correlation (Appendix \ref{SM-pedagog1_details}, 
Fig. \ref{SM-fig_transf_pearson_spearman}). Although correlation
is only one way to 
study relationships between variables, other common metrics also 
provide partial information.

@@

<!--***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*, which we here 
introduce. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure.
<!--***DANr: Maybe move the below two sentences into the new section on Background-->
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A main benefit of a copula approach is 
that it can detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for Q1 below).

The goal of this paper is to explore the potential for 
applications of copulas in ecology. 
<!--***DANr: Maybe emphasize here that we provide some, but not a comprehensive,
intro to copulas - for that go elsewhere. Instead we are trying to determine
to what extent they may be useful in ecology.-->
Our results provide support for the hypothesis that environmental, ecological and evolutionary processes 
generate and transmit nontrivial 
<!--***DANr: nontrivial? what does it mean-->
copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. The work we present here has begun to test this
hypothesis, and we hope our results stimulate further
evaluation of the hypothesis by the research community.

@@

<!--***DANr: Maybe use the words tail dependence more and the word copula
less in the below? Generally the words "copula structure" may be rubbing some
people wrong. Replacement words? What do you actually mean here? Maybe call it
asymmetric dependence, or complex dependence structure?-->
One way we may expect, *a priori*, a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association is visible in the copula
approach, but may not be revealed by standard techniques.

@@

<!--***DANr: Need some caveating here to accommodate the people who slavishly
follow the old maxim that correlation is not causation.-->
A third reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
would suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F suggests that Sp2
is the dominant competitor. The copula approach again reveals which portions of the
distributions are principally responsible for the association between 
variables, and this 
information can then help support or refute alternative causal hypotheses
about which species is driving the dynamics of the other.

@@

# Background on copulas

<!--***DANr: Add a new section above this one called Background on copulas.
It can have the following kinds of stuff (probably not exhaustive, and probably
not in order, and some of this material is currently below):
The definition
The explanation of how ranks relate to estimating the copula parameters
How to estimate carry out fitting (basically, MLE fitting on ranks)
What some of the different copula families look like
What tail dependence actually is
-->
